---
title: "P8105 Data Science I - Homework 2"
author: "Heng Hu (hh2648)"
date: 2025-09-30
output: github_document
---

Load packages

```{r message = FALSE}
library(tidyverse)
library(readxl)
```


## Problem 1

```{r message = FALSE}
# Input and clean pols-month.csv
pols = read_csv(file = "./data/pols-month.csv") |> 
  separate(mon, c("year", "month", "day"), remove = FALSE) |> 
  mutate(year = as.numeric(year),
         month = month(mon, label = TRUE, abbr = FALSE),
         president = ifelse(prez_gop == 1, "rep", "dem")) |> 
  select(-c(mon, day, prez_gop, prez_dem)) |> 
  select(year, month, president, everything()) |> 
  arrange(year, month)

pols
```

The dataset `pols` has ``r ncol(pols)`` variables and ``r nrow(pols)`` records related to the number of national politicians who are democratic or republican. The dataset covers the period from ``r pols[1, "month"]``, ``r pols[1, "year"]`` to ``r pols[nrow(pols), "month"]``, ``r pols[nrow(pols), "year"]``. Variables `year` and `month` represent the time period, while `president` indicates the political party of the president during that time.

```{r message = FALSE}
# Input and clean snp.csv
snp = read_csv(file = "./data/snp.csv") |> 
  mutate(date = as.Date(date, format = "%m/%d/%y"),
         year = year(date),
         year = ifelse(year > 2025, year-100, year),
         month = month(date, label = TRUE, abbr = FALSE)) |> 
  select(year, month, close) |> 
  arrange(year, month)

snp
```

The dataset `snp` has ``r ncol(snp)`` variables and ``r nrow(snp)`` records related to Standard & Poor’s stock market index (S&P). The dataset covers the period from ``r snp[1, "month"]``, ``r snp[1, "year"]`` to ``r snp[nrow(snp), "month"]``, ``r snp[nrow(snp), "year"]``. Variables `year` and `month` represent the time period, while `close` indicates the closing values of the S&P stock index on the associated date.

```{r message = FALSE}
# Input and clean unemployment.csv
unemployment = read_csv(file = "./data/unemployment.csv") |> 
  pivot_longer(cols = (-Year),
               names_to = "month_abb",
               values_to = "ue_rate") |> 
  mutate(year = Year,
         month = match(month_abb, month.abb),
         yr_mon = as.Date(paste(year, month, "01", sep = "-")),
         month = month(yr_mon, label = TRUE, abbr = FALSE)) |> 
  select(year, month, ue_rate) |> 
  arrange(year, month)

unemployment
```

The dataset `unemployment` has ``r ncol(unemployment)`` variables and ``r nrow(unemployment)`` observations recording the unemployment rate. The dataset covers the period from ``r unemployment[1, "month"]``, ``r unemployment[1, "year"]`` to ``r unemployment[nrow(unemployment), "month"]``, ``r unemployment[nrow(unemployment), "year"]``. Variables `year` and `month` represent the time period, while `ue_rate` indicates the unemployment rate during that time.

```{r message = FALSE}
# Join the 3 datasets 
full_dt = pols |> 
  left_join(snp, by = c("year", "month")) |> 
  left_join(unemployment, by = c("year", "month"))

full_dt
```

Since the merged data may be used to analyze the political influence on national-level economic conditions, I applied `left_join` to retain all the observations from the `pols` dataset when merging these 3 datasets. The merged dataset is called `full_dt`. It contains ``r nrow(full_dt)`` observations of ``r ncol(full_dt)`` variables, and covers the period from ``r full_dt[1, "month"]``, ``r full_dt[1, "year"]`` to ``r full_dt[nrow(full_dt), "month"]``, ``r full_dt[nrow(full_dt), "year"]``. Like the previous datasets, variables `year` and `month` represent the time period. `president`, `close`, and `ue_rate` indicate, respectively, the political party of the president, the closing values of the S&P stock index, and unemployment rate during that time.



## Problem 2

```{r}

# Read and clean the Mr. Trash Wheel sheet
mr_trash = read_excel("./data/202509 Trash Wheel Collection Data.xlsx",
                      sheet = "Mr. Trash Wheel",
                      range = cell_cols("A:N"),
                      skip = 1) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(year = as.numeric(year),
         sports_balls = as.integer(round(sports_balls, 0)),
         wheel_name = "Mr. Trash Wheel") |> 
  select(wheel_name, everything())
mr_trash

# Read and clean the Professor Trash Wheel sheet
prof_trash = read_excel("./data/202509 Trash Wheel Collection Data.xlsx",
                      sheet = "Professor Trash Wheel",
                      range = cell_cols("A:M"),
                      skip = 1) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(wheel_name = "Professor Trash Wheel") |> 
  select(wheel_name, everything())

prof_trash

# Read and clean the Gwynns Falls Trash Wheel sheet
gwynns_trash = read_excel("./data/202509 Trash Wheel Collection Data.xlsx",
                      sheet = "Gwynns Falls Trash Wheel",
                      range = cell_cols("A:L"),
                      skip = 1) |> 
  janitor::clean_names() |> 
  filter(!is.na(dumpster)) |> 
  mutate(wheel_name = "Gwynns Falls Trash Wheel") |> 
  select(wheel_name, everything())

gwynns_trash

# Combine these datasets
three_trash_wheel = bind_rows(mr_trash, prof_trash, gwynns_trash)

three_trash_wheel
```

The dataset `mr_trash` has ``r ncol(mr_trash)`` variables and ``r nrow(mr_trash)`` records, with a time span ranging from ``r mr_trash[1, "month"]``, ``r mr_trash[1, "year"]`` to ``r mr_trash[nrow(mr_trash), "month"]``, ``r mr_trash[nrow(mr_trash), "year"]``. The dataset `prof_trash` has ``r ncol(prof_trash)`` variables and ``r nrow(prof_trash)`` records, with a time span ranging from ``r prof_trash[1, "month"]``, ``r prof_trash[1, "year"]`` to ``r prof_trash[nrow(prof_trash), "month"]``, ``r prof_trash[nrow(prof_trash), "year"]``. The dataset `gwynns_trash` has ``r ncol(gwynns_trash)`` variables and ``r nrow(gwynns_trash)`` records, with a time span ranging from ``r gwynns_trash[1, "month"]``, ``r gwynns_trash[1, "year"]`` to ``r gwynns_trash[nrow(gwynns_trash), "month"]``, ``r gwynns_trash[nrow(gwynns_trash), "year"]``. All of them share some common variables listed here. The `wheel_name` indicate the name of trash wheel. `year`, `month`, and `date` represent the time period. `weight_tons` and `volume_cubic_yards` are the total weight and volume of trash that the trash wheel collected. `homes_powered` shows the number of homes powered by the energy generated from burning collected trash. Other variables like ``r  names(three_trash_wheel |> select(plastic_bottles:sports_balls))`` show the count numbers of different types of trash. However, not all of them are present in every dataset.

The combined dataset `three_trash_wheel` has ``r ncol(three_trash_wheel)`` variables and ``r nrow(three_trash_wheel)`` records. It contains data collected from ``r unique(three_trash_wheel$wheel_name)`` and all the variables listed in previous paragraph.

Based on the current data, the total weight of trash collected by Professor Trash Wheel is ``r sum(prof_trash |> select(weight_tons))`` tons. The total number of cigarette butts collected by Gwynnda in June of 2022 was ``r sprintf("%d", sum(gwynns_trash |> filter(year == 2022 & month == "June") |> select(cigarette_butts)))``.
